---
title: "Santander_Case"
author: "Sahana Ramakrishnan"
date: "2024-11-21"
output:
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(summarytools)
library(gridExtra)
```

## Santander Bank Case Study

### Loading data
```{r}
#train and test data
train <- read_csv("D:/NewEng/Fall - Winter 2024/Santander case study/data set/train.csv")
test <- read_csv("D:/NewEng/Fall - Winter 2024/Santander case study/data set/test.csv")

head(train)
head(test)

```
- Loaded Santander data set
- Checking the data using head function


### Merging the data
```{r}

# Adding a column to differentiate train and test
train <- train %>% mutate(dataset = "train")
test <- test %>% mutate(dataset = "test")

# combining data for unified EDA
combined_data <- bind_rows(train, test)

```

- A new column is added to differentiate between the training and test datasets which are then combined using the bind_rows() function for unified EDA.
- This allows seamless exploration of patterns across both datasets while keeping them distinguishable

### Performing EDA
```{r}
# Distribution of target column var.
table(train$target)
target_distribution <- ggplot(train, aes(x = factor(target))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Target variable", x = "Target", y = "Count") +
  theme_minimal()

target_distribution

```

- The majority class: target = 0, significantly outweighs the minority class: target = 1, showing severe class imbalance in the dataset.

- This imbalance suggests that specialized techniques like resampling or adjusting class weights may be required to ensure model performance is not biased toward the majority class

### Feature Analysis
```{r}
# summary of numerical predictors
summary(train %>% select(starts_with("var_")))

# corr. matrix of some predictors
predictor_subset <- train %>% select(var_0:var_10)
correlation_matrix <- cor(predictor_subset)
correlation_matrix

```

- Summary statistics are calculated for the numerical predictors to understand their range and distribution.
- A correlation matrix of selected predictors: var_0 to var_10 is generated to identify relationships and dependencies between features.

### plotting predictors
```{r}
# scatterplot of 2 predictors
scatter_plot <- ggplot(train, aes(x = var_0, y = var_1, color = factor(target))) +
  geom_point(alpha = 0.6) +
  labs(title = "Scatterplot of var_0 and var_1", x = "var_0", y = "var_1") +
  theme_minimal()

scatter_plot

# Distribution of single predictor
predictor_distribution <- ggplot(train, aes(x = var_2)) +
  geom_density(fill = "purple", alpha = 0.6) +
  labs(title = "Density Plot of var_2", x = "var_2", y = "Density") +
  theme_minimal()

predictor_distribution

```


- The scatterplot indicates the target classes 0 and 1 are not well-separated for the features var_0 and var_1 suggesting limited predictive power of these individual features.

- The overlapping points across classes highlight the need for additional feature engineering or combining multiple features to improve class distinction.

- The density plot of var_2 shows a unimodal distribution with a peak between 9 and 10 indicating that most values of var_2 are concentrated within this range.

- There is no visible outliers or anomalies in the density plot showing var_2 a stable predictor in the dataset
